---
title: "ADGRs in lymphatic endothelial cells in scRNA-seq datasets"
author: "Heli Xu"
format: html
editor: visual
---

## Dataset1: postMI mice heart non-CM

Publication source <https://elifesciences.org/articles/43882>.

### Markers used to define LEC

The paper did not include cell type information in the dataset, so we will use established markers to define endothelial cells (EC) and lymphatic endothelial cells (LEC).

EC markers used include: Pecam1, Icam2, Cdh5 (code or image to be added)

```{r}
#dependencies 
```

Based on the plot, we'll define cluster 2, 7, 9, 12, 15, 18, 19 as EC.

Among them, we'll use LEC lineage markers to identify possible LEC populations. Markers used include: Pdpn, Flt4, Prox1, Ackr4, Msr1, Fcgr2b, Lyve1, Fth1

```{r}
load("data/MI_EC_normalized.rdata")
##this is EC subset without changing idents to "sample"!

VlnPlot(TIP_MI_EC, 
        features = c("Pdpn","Flt4","Prox1","Ackr4","Msr1","Fcgr2b", 
                     "Lyve1", "Fth1"),
        pt.size = 0)

```

It looks like the cluster 12, 15, 18 are likely to be LEC. Firstly, we'll plot all the ADGRs in EC, keeping the LEC clusters in mind. 

```{r}
ADGR_list <- c("Adgra1", "Adgra2","Adgra3", "Adgrb1","Adgrb2", "Adgrb3",
               "Adgrc3","Adgrc1", "Adgrc2", "Adgrd1", "Adgrd2", "Adgre1",
               "Adgre2", "Adgre3", "Adgre4", "Adgre5", "Adgrf1", "Adgrf2",
               "Adgrf3", "Adgrf4","Adgrf5","Adgrg1","Adgrg2","Adgrg3",
               "Adgrg4", "Adgrg5", "Adgrg6", "Adgrg7","Adgrl1", "Adgrl2",
               "Adgrl3", "Adgrl4", "Adgrv1")

FeaturePlot(TIP_MI_EC,
            reduction = "tsne",
            features= ADGR_list,
            sort.cell = TRUE,
            min.cutoff = 'q10',
            label = TRUE)

